Predicting Fire Risk: \\ An Evaluation of the Fire Danger Index

### 2016-01-22 paper.tex
spell check

In Australia, bushfires are one of the most common natural disasters impacting lives and property. To minimise the damage from fire occurence, management agencies often allocate resources according to where fires are expected to occur and where fire behaviour could be more severe  \citep{padilla11, wotton05}. In order to plan effectively and manage risks resulting from bushfire, the measures used to predict fire occurrence and severity must be reliable and accurate.

@@

I construct a dataset of bushfire ignition and metereological variables such that predicting fire ignition becomes a binary classification problem. Then I apply both some common machine learning techniques: classification trees and random forests and compare their performance to logistic regression models which use non-parametric smoothing functions. The dataset suffers from severe class imbalance, the number of fires compared to number of non fires is tiny. Accordingly, models are evaluated based on balance accuracy measures and reciever operating characteristic plots.

@@

While the FFDI may be a good measure of fire intensity once ignition occurs, it is a poor predictor of fire ignition. Predictor performance is improved with the inclusion of the FFDI input variables, grassland curing, vapour pressure at 9am and precipitation. Specfically of these variables, grassland curing is the most important variable in determining probability of ignition.

@@

McArthur’s FFDI originated as tabulated indices and were subsequently reduced to equations by \citet{noble80} which in turn simplified into the formulation of the FFDI given by equation (\ref{eq FFDI}).

@@

There is extensive research on predicting fire behaviour using meteorological and topographical information given ignition occurence but comparatively little also on  predicting fire occurrence and on evaluating the McArthur index.

@@

The metereological data set comes from a climate simulation of the Murrary Darling Basin using the Weather Research and Forecasting (WFR) model. The WRF Modeling system has been developed by the National Center for Atmospheric Research (NCAR), the National Oceanic and Atmospheric Administrations Centers for Environmental Predicion (NCEP) and the other weather research institutions in the United States. \citet{evans10} used the Advanced Research WFR version 3 \citep{skamarock08} to simulate weather for southeastern Australia at a 10km horizontal grid over 24 years from 1985.

@@

The FFDI data is created with the metereological variables from the WRF model. As per \citet{clarkeevans13} the spatial distribution of the FFDI has also been evaluated against observations with success  \citep{sanabria13}. Clarke and Evans discuss some heterogenieties with modelled windspeed due to changing technology used to measure this variable mostly evident in rural areas before the 1990s resulting in an underestimation of average FFDI values by approximately 5 percent for this period. The fire occurence dataset begins in 2000 and as a result these windspeed errors do not impact my analysis.

The WRF model \citep{evans10} is deemed to be an accurate representation of Australian weather on this scale and is a useful tool for modelling fire weather over Australia \citep{clarke2013}. The paper uses the FFDI, maximum temperature (Kelvin), wind speed (m/s), relative humidity (\%) and the drought factor calculated by \citet{clarke2013}  from this model. All other variables used in this paper are matched to the 10km gridpoints in the meteorological dataset.

%Graphs, need to redo better
\begin{figure}[h]
	\centering 
%	\includegraphics[width=1\columnwidth]{figures/Data/silvan.pdf} 
	\caption{Time series of meterological varibles for Silvan (-37.82, 145.42)} 
	\label{fig:silv} 
\end{figure}

@@

Combination of these variables as per equation (\ref{eq FFDI}) yields the FFDI which clearly exhibits seasonality. Note the spike in summer 2009, correspinding to the Black Saturday bushfires.

@@

The plots for districts 15 and 20 are missing due to there not being enough datapoints in each fire dating rating level to produce the 2D densities.

\subsubsection{Grassland Curing}
The level of dying or senescence of plant material caused by seasonal weather patterns is refered to as grassland curing. Grassland curing ranges between 0 and 100, where 0 means the grass is green with high moisture content and 100 means the grass is at its driest. Curing has a significant impact on fire behaviour in grasslands, particularly the probability of ignition and subsequent rate of spread \citep{cheney08}.

@@

Figure \ref{fig:cur_lp} shows a level plot produced using the rasterVis package \citep{rastervis}. We observe grassland curing values across the state for the week beginning 25th January 2009. The horizontal margin shows the mean curing value for each longitude and the vertical margin shows the mean curing for each lattitude. The plot clearly shows the difference in curing values across the state, in particular the difference between the east and west of the state. This is consistent with time series graphs of curing.

\subsubsection{Vapour Pressure}
Vapour pressure, in meteorology, describes the partial pressure of water vapour in the atmosphere, it is directly related to the number of water vapour molecules in the air.  \citet{harris14} find there is a significant relationship between fire activity and vapour pressure among other climate variables. I investigate whether vapour pressure at 9am can improve prediction of fire occurence.

@@

Table \ref{table:correl} shows the cross correlaton matrix for the variables used to predict probability of ignition. High correlation can pose a problem in regression models and to a lesser extent machine learning methods. Variables of concern are temperature, drought factor, relative humidity and FFDI and  which have relatively strong correlation coefficients.

\subsection{Response Variable: Fire Occurence}

@@

Figure \ref{fig:firemap} shows the locations of the fires in the dataset on the map of Victoria. The most noticable thing in Figure \ref{fig:firemap} is the sparsity of fires in the north-east and north west of the state and around Melbourne. This is due to the division of fire fighting organisations in Victoria. The sparse areas correspond to National Parks where Parks Victoria or the Department of Environment, Land, Water and Planning (DELWP) coordinate fire fighting activities, so many of the fires occuring in these areas are not included in this dataset.

@@

Figure \ref{fig:tw_fire} is constructed in the same manner as Figure \ref{fig:tw_fdr} but instead of grouping by the fire dange rating, I group by Fire/No Fire. Windspeed and temperature play the largest role in determining the FFDI yet there is no discernable difference between the distributions of the two variables when there is a fire compared to no fires. This is the first indicator that the FFDI alone may not be a good predictor of probability of ignition.

@@

Plotting the 2D densities of the drought factor and relative humidity in Figure \ref{fig:dh_fire}, ranges of relative humidty don't appear to vary on fire days compared to not fire days but in districts 4, 5, 6, 7, 8, 9, 10, 11, 13, 14 and 24 there is a clear difference in the densities of fire compared to no fire. Small drought factors are mostly associated with non-fire days. In districts 2, 12, 15, 16, 17, 18, 20 and 22 do not have large differences in the two densities, these districts generally have high drought factor and low relative humidity.

@@

Assembling the datasets began with the meterological data from the WRF model. This data came in NetCDF form \citep{ncdf}. The dataset spanned all of Eastern Australian (Queensland, New South Wales \& Victoria) with daily observations over a 25 year period. This area, divided into 10km grids, contained 28675 gridpoints each of which had longitude, lattitude and weather variable values (temperature, wind speed, relative humidty, FFDI and drought factor) associated with it. Once the time component is taken into account, the original dataset contained over 200 million data points (28675 $\times$ 365 $\times$ 25 ).

The first step was to reduce the dataset to a more manageable size. I converted the NCDF file to a datatable using the raster and data.table packages in R \citep{raster, datatable}. I used ESRI shapefiles of Australian Geographical Classification boundaries \citep{ABS1259} to sort the gridpoints into states and selected only gridpoints in Victoria. The datasets intersect between 2000 and 2009, subsetting the WFR data reduces the size of the dataset to just over 8 million data points.

@@

From the fire occurence dataset this study used the date and coordinates of the fire. The coordinates were given in the Australian Map Grid (AMG66), a coordinate reference system derived from a Universal Transverse Mercator(UMT) projection of latitudes and longitudes on the Australian Geodetic Datum (AGD) \citep{featherstone96}. To merge the meterological data set and the fire occurence dataset I needed to convert from this UMT projection to longitude and lattitude using the sp package in R \citep{sp08}.

Once the coordinates of fires were converted to the same projection I matched each fire coordinate to its nearest neighbour from the meterological dataset. Once matched it was straightforward to merge the two datasets based on grid point and date \citep{datatable}.

The different CFA districts and regions were added into the dataset using the raster package and another ESRI shapefile containing the district boundaries \citep{raster}. 
Digital boundaries for the CFA area compared to MFB or DELWP response areas were unavailable, as such an alternate method to subset geographical area was used. I made the assumption that if the CFA had not attended a bushfire betwen 2000 and 2014 within a 5km region of the gridpoint, then the gridpoint was unlikely to be in a CFA area. Accordingly any grids that had no fire occur in them since 2000 were excluded from the dataset.

The three additional variables (grassland curing, vapour pressure and precipitation) came as raster files each of higher resolutions than the original 10km grid. In all three sets we have some missing values or faulty readings due to technological fault. To minimise the impact of these for each for each gridpoint in the meteorological dataset I find the pixels contained in each gridpoint and take the median value for each time period. The functionality to combine spatial points and raster layers is provided by the raster package \citep{raster}. For any remaining missing values I make the assumption that cloud cover or faulty barometer readings are sufficiently random such that missing values will not bias results.

@@

As a result the logistic model becomes
\begin{equation}
\label{eq logits}
\Pr(Y=1|X)=\frac{\exp(\sum_{k=1}^K f_k(x_k))}{1+\exp(\sum_{k=1}^K f_k(x_k)}
\end{equation}
where $Y$ denomtes fire occurence, $x_k$ denotes the $k^{th}$ predictor and each $f_k$ represents a penalized regression spline.

@@

The task of finding the single best predictor is achieved using recursive partitioning of the data and can be summarized as two major steps:
\begin{enumerate}
	\item For each predictor multiple cut points are evaluated and the quality of the split is based on the Gini index. Equation (\ref{eq gini}) gives the Gini index for a binary variable:
	\begin{equation}
	\label{eq gini}
	G=\sum_{k=1}^{K} \hat{p}_{mk} (1-\hat{p}_{mk})
	\end{equation}
	where $\hat{p}$ is the estimated probability of a fire at that node. It is also possible to use the cross entropy \citep{james13} which like the Gini index takes on a small value if the node is pure. 
	For each predictor the best cut point is chosen. 
	\item The best split overall is chosen:  For each predictor and its corresponding ‘best’ split, the overall best split is chosen again based on Gini index. 
\end{enumerate}

@@

In the context of fire occurence, even during the fire season in Victoria instances of bushfire are not that common. For any given day in the dataset at most 4.5\% of the grids experience fire, some days it is possible to have no fires at all. Intuitively, before looking at the data we can expect the proportion of fire days is much smaller than that of non fire days.

Restricting the dataset to the Victorian fire season (October to March) marginally improves the imbalance problem; nevertheless the proportion of fire days of total days for each spatial point is 0.0053. It is safe to say that this situation is one where the degree of imbalance is problematic as a classifier which always assumes `no fire' has an expected error rate of 0.5\% leaving very little room for a more complicated classifer to improve upon.

There are several ways to address the class imbalance problem in machine learning: over sampling, undersampling, using different splitting criteria \citep{japkowicz00}.

@@

Undersampling the no fire class addresses the problem of imbalanced classes but I also use it to reduce the serial and spatial correlation that is present in the dataset by using a ~near-maximin' sampling method thereby ensuring all the datapoints in the sample are separate from each from each other in space and time to reduce the impact of correlation.

@@

I also report sensitivity (correctly classified fires dividied by total number of actual fires) and specificity (correctly identified non-fires divided by total number of non-fires), balanced accuracy and kappa.

@@

where $P_0$ is the total agreement probability and $P_c$ is the “agreement” probability which is due to chance ( see \cite{kappa} for definitions )

@@

A likelihood ratio test rejects the null that temperature, relative humidity, windspeed and drought factor are jointly insignificant providing further evidence that the FFDI is not using the meteorological information efficiently and there are relationships between fire occurence and these variables that the FFDI is not capturing.

@@

The drought factor has an almost linear impact on the log-odds ratio when it is inbetween 0 and 6. When drought factor is between 0 and 6 it has a negative effect on log-odds but at a decreasing rate. When the drought factor it between 6 and 9 it has a positive impact on the log-odds ratio, after 9 the effect because negative again.

Relative humidty appears to have a quadratic relationship with the log-odds ratio. Relative humity only has positive effect on the log-odds ration when it is inbetween 30 and 70\%. This function makes sense intuitively as when relative humidty is very high there is too much moisture in the air (or it is raining) preventing ignition. When relative humidity is very low it does not appear to have a statistically significant impact on the log-odd ratio. The middle values for relative humidty would correspond to warm, dry days, ideal ignition conditions.

@@

%GAM plots, not sure if this is exactly splines... 
\begin{figure}[h]
\null\vfill\noindent
%\includegraphics[page=1,width=.48\linewidth,height=.45\textheight,keepaspectratio]{figures/gam_pl.pdf}
\hfill
%\includegraphics[page=2,width=.48\linewidth,height=.45\textheight,keepaspectratio]{figures/gam_pl.pdf}
\vfill
%\includegraphics[page=3,width=.48\linewidth,height=.45\textheight,keepaspectratio]{figures/gam_pl.pdf}
\hfill
%\includegraphics[page=4,width=.48\linewidth,height=.45\textheight,keepaspectratio]{figures/gam_pl.pdf}
\vfill
%\includegraphics[page=5,width=.48\linewidth,height=.45\textheight,keepaspectratio]{figures/gam_pl.pdf}
\hfill
%\includegraphics[page=6,width=.48\linewidth,height=.45\textheight,keepaspectratio]{figures/gam_pl.pdf}
\vfill
\null
	\caption{Plots of smoothing functions in Model 4 with 95\% pointwise confident interval} 
	\label{fig:gam_pl} 
\end{figure}

@@

The classification tree results reaffirm the results of the spline models in figure \ref{fig:gam_pl} which suggested the effect of curing and drought factor on probability of ignition is not constant at each level of the variable. The bands of values chosen by the tree as having higher probability of ignition vaguely match the shape of the fitted spline functions in Model 4. The tree selecting ranges of variables for which probability of ignition changes suggests there is a  non-linear relationship between some of the explanatory variables and probabiity of ignition.

@@

The classification trees consistently place  curing value and drought factor high up in the tree. This makes intuitive sense as both these variables are related to the fuel and the dryness of the available fuel which potentially is more important that the air temperature. Temperature is expected to contribute to probability of ignition but in the longer term. Prolonged high temperatures lead to high curing (drier grass) and higher drought factor level, but according to the classifcation tree, temperature on the particular day is irrelevant.

@@

FFDI, temperature, vapour pressure, relative humidity and windspeed all have close values of importance. The FFDI is second most important in determining probability of fire ignition which seems counterintuitive given it's poor perfomance in the classification trees and logistic regression.

@@

Obviously there's more to fire occurence than weather conditions, this study leaves out the essential human component of fire ignition. We see in Figure \ref{fig:firemap} that fire activity corresponds with population density and the major roads in Victoria. The best model in this thesis still only has balanced accuracy of .64 which is a fairly poor model but a more accurate model would indicate overfitting or something else has gone awry.  Weather conditions are important to an extent but we expect them to have greater impact on spread and fire behaviour once ignition has occured rather than ignition itself.

@@

The single classification trees placed FFDI as very low split which we can interpret as the FFDI not being very helpful in classifying fire occurence. The random forest ranked the importance of the FFDI second highest but given the correlation between the FFDI and its inputs this result is unreliable.

The logistic model with splines, classification trees and random forest all agreed on the importance of grassland curing in predicting fire ignition. The statistical significance of the spline function of curing in the logistic-spline model, the output of the classification trees consistently placing curing as the highest split and the variable importance ouput from the random forest. While all three additional variables improved accuracy of the models, grassland curing was clearly superior.

@@

The interesting result is the similarity in perfomance of the splines model and the random forest. Using splines to allow a non-linear relationship between the log-odds ratio and the variables while random forests typically capture interactions between variables. Both these mathods led to a large improvement in the predictive perfomance.

@@

>>>>

In Australia, bushfires are one of the most common natural disasters impacting lives and property. To minimise the damage from fire occurrence, management agencies often allocate resources according to where fires are expected to occur and where fire behaviour could be more severe  \citep{padilla11, wotton05}. In order to plan effectively and manage risks resulting from bushfire, the measures used to predict fire occurrence and severity must be reliable and accurate.

@@

I construct a dataset of bushfire ignition and meteorological variables such that predicting fire ignition becomes a binary classification problem. Then I apply both some common machine learning techniques: classification trees and random forests and compare their performance to logistic regression models which use non-parametric smoothing functions. The dataset suffers from severe class imbalance, the number of fires compared to number of non fires is tiny. Accordingly, models are evaluated based on balance accuracy measures and receiver operating characteristic plots.

@@

While the FFDI may be a good measure of fire intensity once ignition occurs, it is a poor predictor of fire ignition. Predictor performance is improved with the inclusion of the FFDI input variables, grassland curing, vapour pressure at 9am and precipitation. Specifically of these variables, grassland curing is the most important variable in determining probability of ignition.

@@

McArthur's FFDI originated as tabulated indices and were subsequently reduced to equations by \citet{noble80} which in turn simplified into the formulation of the FFDI given by equation (\ref{eq FFDI}).

@@

There is extensive research on predicting fire behaviour using meteorological and topographical information given ignition occurrence but comparatively little also on  predicting fire occurrence and on evaluating the McArthur index.

@@

The meteorological data set comes from a climate simulation of the Murray Darling Basin using the Weather Research and Forecasting (WFR) model. The WRF Modeling system has been developed by the National Center for Atmospheric Research (NCAR), the National Oceanic and Atmospheric Administrations Centers for Environmental Prediction (NCEP) and the other weather research institutions in the United States. \citet{evans10} used the Advanced Research WFR version 3 \citep{skamarock08} to simulate weather for southeastern Australia at a 10km horizontal grid over 24 years from 1985.

@@

The FFDI data is created with the meteorological variables from the WRF model. As per \citet{clarkeevans13} the spatial distribution of the FFDI has also been evaluated against observations with success  \citep{sanabria13}. Clarke and Evans discuss some heterogeneities with modelled windspeed due to changing technology used to measure this variable mostly evident in rural areas before the 1990s resulting in an underestimation of average FFDI values by approximately 5 percent for this period. The fire occurrence dataset begins in 2000 and as a result these windspeed errors do not impact my analysis.

The WRF model \citep{evans10} is deemed to be an accurate representation of Australian weather on this scale and is a useful tool for modelling fire weather over Australia \citep{clarke2013}. The paper uses the FFDI, maximum temperature (Kelvin), wind speed (m/s), relative humidity (\%) and the drought factor calculated by \citet{clarke2013}  from this model. All other variables used in this paper are matched to the 10km grid points in the meteorological dataset.

%Graphs, need to redo better
\begin{figure}[h]
	\centering 
%	\includegraphics[width=1\columnwidth]{figures/Data/silvan.pdf} 
	\caption{Time series of meteorological variables for Silvan (-37.82, 145.42)} 
	\label{fig:silv} 
\end{figure}

@@

Combination of these variables as per equation (\ref{eq FFDI}) yields the FFDI which clearly exhibits seasonality. Note the spike in summer 2009, corresponding to the Black Saturday bushfires.

@@

The plots for districts 15 and 20 are missing due to there not being enough data points in each fire dating rating level to produce the 2D densities.

\subsubsection{Grassland Curing}
The level of dying or senescence of plant material caused by seasonal weather patterns is referred to as grassland curing. Grassland curing ranges between 0 and 100, where 0 means the grass is green with high moisture content and 100 means the grass is at its driest. Curing has a significant impact on fire behaviour in grasslands, particularly the probability of ignition and subsequent rate of spread \citep{cheney08}.

@@

Figure \ref{fig:cur_lp} shows a level plot produced using the rasterVis package \citep{rastervis}. We observe grassland curing values across the state for the week beginning 25th January 2009. The horizontal margin shows the mean curing value for each longitude and the vertical margin shows the mean curing for each latitude. The plot clearly shows the difference in curing values across the state, in particular the difference between the east and west of the state. This is consistent with time series graphs of curing.

\subsubsection{Vapour Pressure}
Vapour pressure, in meteorology, describes the partial pressure of water vapour in the atmosphere, it is directly related to the number of water vapour molecules in the air.  \citet{harris14} find there is a significant relationship between fire activity and vapour pressure among other climate variables. I investigate whether vapour pressure at 9am can improve prediction of fire occurrence.

@@

Table \ref{table:correl} shows the cross correlation matrix for the variables used to predict probability of ignition. High correlation can pose a problem in regression models and to a lesser extent machine learning methods. Variables of concern are temperature, drought factor, relative humidity and FFDI and  which have relatively strong correlation coefficients.

\subsection{Response Variable: Fire Occurrence}

@@

Figure \ref{fig:firemap} shows the locations of the fires in the dataset on the map of Victoria. The most noticeable thing in Figure \ref{fig:firemap} is the sparsity of fires in the north-east and north west of the state and around Melbourne. This is due to the division of fire fighting organisations in Victoria. The sparse areas correspond to National Parks where Parks Victoria or the Department of Environment, Land, Water and Planning (DELWP) coordinate fire fighting activities, so many of the fires occurring in these areas are not included in this dataset.

@@

Figure \ref{fig:tw_fire} is constructed in the same manner as Figure \ref{fig:tw_fdr} but instead of grouping by the fire danger rating, I group by Fire/No Fire. Windspeed and temperature play the largest role in determining the FFDI yet there is no discernible difference between the distributions of the two variables when there is a fire compared to no fires. This is the first indicator that the FFDI alone may not be a good predictor of probability of ignition.

@@

Plotting the 2D densities of the drought factor and relative humidity in Figure \ref{fig:dh_fire}, ranges of relative humidity don't appear to vary on fire days compared to not fire days but in districts 4, 5, 6, 7, 8, 9, 10, 11, 13, 14 and 24 there is a clear difference in the densities of fire compared to no fire. Small drought factors are mostly associated with non-fire days. In districts 2, 12, 15, 16, 17, 18, 20 and 22 do not have large differences in the two densities, these districts generally have high drought factor and low relative humidity.

@@

Assembling the datasets began with the meteorological data from the WRF model. This data came in NetCDF form \citep{ncdf}. The dataset spanned all of Eastern Australian (Queensland, New South Wales \& Victoria) with daily observations over a 25 year period. This area, divided into 10km grids, contained 28675 grid points each of which had longitude, latitude and weather variable values (temperature, wind speed, relative humidity, FFDI and drought factor) associated with it. Once the time component is taken into account, the original dataset contained over 200 million data points (28675 $\times$ 365 $\times$ 25 ).

The first step was to reduce the dataset to a more manageable size. I converted the NCDF file to a data table using the raster and data.table packages in R \citep{raster, datatable}. I used ESRI shapefiles of Australian Geographical Classification boundaries \citep{ABS1259} to sort the grid points into states and selected only grid points in Victoria. The datasets intersect between 2000 and 2009, subsetting the WFR data reduces the size of the dataset to just over 8 million data points.

@@

From the fire occurrence dataset this study used the date and coordinates of the fire. The coordinates were given in the Australian Map Grid (AMG66), a coordinate reference system derived from a Universal Transverse Mercator(UMT) projection of latitudes and longitudes on the Australian Geodetic Datum (AGD) \citep{featherstone96}. To merge the meteorological data set and the fire occurrence dataset I needed to convert from this UMT projection to longitude and latitude using the sp package in R \citep{sp08}.

Once the coordinates of fires were converted to the same projection I matched each fire coordinate to its nearest neighbour from the meteorological dataset. Once matched it was straightforward to merge the two datasets based on grid point and date \citep{datatable}.

The different CFA districts and regions were added into the dataset using the raster package and another ESRI shapefile containing the district boundaries \citep{raster}. 
Digital boundaries for the CFA area compared to MFB or DELWP response areas were unavailable, as such an alternate method to subset geographical area was used. I made the assumption that if the CFA had not attended a bushfire between 2000 and 2014 within a 5km region of the grid point, then the grid point was unlikely to be in a CFA area. Accordingly any grids that had no fire occur in them since 2000 were excluded from the dataset.

The three additional variables (grassland curing, vapour pressure and precipitation) came as raster files each of higher resolutions than the original 10km grid. In all three sets we have some missing values or faulty readings due to technological fault. To minimise the impact of these for each for each grid point in the meteorological dataset I find the pixels contained in each grid point and take the median value for each time period. The functionality to combine spatial points and raster layers is provided by the raster package \citep{raster}. For any remaining missing values I make the assumption that cloud cover or faulty barometer readings are sufficiently random such that missing values will not bias results.

@@

As a result the logistic model becomes
\begin{equation}
\label{eq logits}
\Pr(Y=1|X)=\frac{\exp(\sum_{k=1}^K f_k(x_k))}{1+\exp(\sum_{k=1}^K f_k(x_k)}
\end{equation}
where $Y$ denotes fire occurrence, $x_k$ denotes the $k^{th}$ predictor and each $f_k$ represents a penalized regression spline.

@@

The task of finding the single best predictor is achieved using recursive partitioning of the data and can be summarized as two major steps:
\begin{enumerate}
	\item For each predictor multiple cut points are evaluated and the quality of the split is based on the Gini index. Equation (\ref{eq gini}) gives the Gini index for a binary variable:
	\begin{equation}
	\label{eq gini}
	G=\sum_{k=1}^{K} \hat{p}_{mk} (1-\hat{p}_{mk})
	\end{equation}
	where $\hat{p}$ is the estimated probability of a fire at that node. It is also possible to use the cross entropy \citep{james13} which like the Gini index takes on a small value if the node is pure. 
	For each predictor the best cut point is chosen. 
	\item The best split overall is chosen:  For each predictor and its corresponding split, the overall best split is chosen again based on Gini index. 
\end{enumerate}

@@

In the context of fire occurrence, even during the fire season in Victoria instances of bushfire are not that common. For any given day in the dataset at most 4.5\% of the grids experience fire, some days it is possible to have no fires at all. Intuitively, before looking at the data we can expect the proportion of fire days is much smaller than that of non fire days.

Restricting the dataset to the Victorian fire season (October to March) marginally improves the imbalance problem; nevertheless the proportion of fire days of total days for each spatial point is 0.0053. It is safe to say that this situation is one where the degree of imbalance is problematic as a classifier which always assumes `no fire' has an expected error rate of 0.5\% leaving very little room for a more complicated classifier to improve upon.

There are several ways to address the class imbalance problem in machine learning: over sampling, under-sampling, using different splitting criteria \citep{japkowicz00}.

@@

Under-sampling the no fire class addresses the problem of imbalanced classes but I also use it to reduce the serial and spatial correlation that is present in the dataset by using a ~near-maximin' sampling method thereby ensuring all the data points in the sample are separate from each from each other in space and time to reduce the impact of correlation.

@@

I also report sensitivity (correctly classified fires divided by total number of actual fires) and specificity (correctly identified non-fires divided by total number of non-fires), balanced accuracy and kappa.

@@

where $P_0$ is the total agreement probability and $P_c$ is the Agreement probability which is due to chance (see \cite{kappa} for definitions).

@@

A likelihood ratio test rejects the null that temperature, relative humidity, windspeed and drought factor are jointly insignificant providing further evidence that the FFDI is not using the meteorological information efficiently and there are relationships between fire occurrence and these variables that the FFDI is not capturing.

@@

The drought factor has an almost linear impact on the log-odds ratio when it is in between 0 and 6. When drought factor is between 0 and 6 it has a negative effect on log-odds but at a decreasing rate. When the drought factor it between 6 and 9 it has a positive impact on the log-odds ratio, after 9 the effect because negative again.

Relative humidity appears to have a quadratic relationship with the log-odds ratio. Relative humidity only has positive effect on the log-odds ration when it is in between 30 and 70\%. This function makes sense intuitively as when relative humidity is very high there is too much moisture in the air (or it is raining) preventing ignition. When relative humidity is very low it does not appear to have a statistically significant impact on the log-odd ratio. The middle values for relative humidity would correspond to warm, dry days, ideal ignition conditions.

@@

%GAM plots, not sure if this is exactly splines... 
\begin{figure}[h]
\null\vfill\noindent
%\includegraphics[page=1,width=.48\linewidth,height=.45\textheight,keepaspectratio]{figures/gam_pl.pdf}
\hfill
%\includegraphics[page=2,width=.48\linewidth,height=.45\textheight,keepaspectratio]{figures/gam_pl.pdf}
\vfill
%\includegraphics[page=3,width=.48\linewidth,height=.45\textheight,keepaspectratio]{figures/gam_pl.pdf}
\hfill
%\includegraphics[page=4,width=.48\linewidth,height=.45\textheight,keepaspectratio]{figures/gam_pl.pdf}
\vfill
%\includegraphics[page=5,width=.48\linewidth,height=.45\textheight,keepaspectratio]{figures/gam_pl.pdf}
\hfill
%\includegraphics[page=6,width=.48\linewidth,height=.45\textheight,keepaspectratio]{figures/gam_pl.pdf}
\vfill
\null
	\caption{Plots of smoothing functions in Model 4 with 95\% point-wise confident interval} 
	\label{fig:gam_pl} 
\end{figure}

@@

The classification tree results reaffirm the results of the spline models in figure \ref{fig:gam_pl} which suggested the effect of curing and drought factor on probability of ignition is not constant at each level of the variable. The bands of values chosen by the tree as having higher probability of ignition vaguely match the shape of the fitted spline functions in Model 4. The tree selecting ranges of variables for which probability of ignition changes suggests there is a  non-linear relationship between some of the explanatory variables and probability of ignition.

@@

The classification trees consistently place  curing value and drought factor high up in the tree. This makes intuitive sense as both these variables are related to the fuel and the dryness of the available fuel which potentially is more important that the air temperature. Temperature is expected to contribute to probability of ignition but in the longer term. Prolonged high temperatures lead to high curing (drier grass) and higher drought factor level, but according to the classification tree, temperature on the particular day is irrelevant.

@@

FFDI, temperature, vapour pressure, relative humidity and windspeed all have close values of importance. The FFDI is second most important in determining probability of fire ignition which seems counterintuitive given it's poor performance in the classification trees and logistic regression.

@@

Obviously there's more to fire occurrence than weather conditions, this study leaves out the essential human component of fire ignition. We see in Figure \ref{fig:firemap} that fire activity corresponds with population density and the major roads in Victoria. The best model in this thesis still only has balanced accuracy of .64 which is a fairly poor model but a more accurate model would indicate overfitting or something else has gone awry.  Weather conditions are important to an extent but we expect them to have greater impact on spread and fire behaviour once ignition has occurred rather than ignition itself.

@@

The single classification trees placed FFDI as very low split which we can interpret as the FFDI not being very helpful in classifying fire occurrence. The random forest ranked the importance of the FFDI second highest but given the correlation between the FFDI and its inputs this result is unreliable.

The logistic model with splines, classification trees and random forest all agreed on the importance of grassland curing in predicting fire ignition. The statistical significance of the spline function of curing in the logistic-spline model, the output of the classification trees consistently placing curing as the highest split and the variable importance output from the random forest. While all three additional variables improved accuracy of the models, grassland curing was clearly superior.

@@

The interesting result is the similarity in performance of the splines model and the random forest. Using splines to allow a non-linear relationship between the log-odds ratio and the variables while random forests typically capture interactions between variables. Both these methods led to a large improvement in the predictive performance.

@@

\section*{Acknowledgment}
\textit{No one who achieves success does so without acknowledging the help of others. The wise and confident acknowledge this help with gratitude. -Alfred North Whitehead
	}

I am indebted to my indefatigable supervisors Rob and Di, thank you for your endless patience.

To those who I adopted as my pseudo-supervisors Souhaib Ben-Taieb, Klaus Ackerman, Simon Angus, your help and advice was invaluable.

To my biggest supporters Earo Wang \& Christien Blencowe without your encouragement, your help and Earo's endless of chocolate this year would have been impossible.

Special thanks to Sarah Harris (Earth \& Atmosphereic Science, Monash), Hamish Clarke (UNSW), Stuart Matthews (Rural Fire Service, NSW), Musa Kilinc \& Danielle Wright (Country Fire Authority), who all contributed with data and helpful advice.

To my long suffering honours cohort, I apologise for the tears and for boring you with my data struggles, you have been the the best team to be a part of.
